---
title: "disease_map_sbc"
output:
  html_document: default
  pdf_document: default
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  include = TRUE,  cache = FALSE,  collapse = TRUE,  echo = TRUE,
  message = FALSE, tidy = FALSE,  warning = FALSE,   comment = "  ",
  dev = "png", dev.args = list(bg = '#FFFFF8'), dpi = 300,
  fig.align = "center",  fig.width = 7,  fig.asp = 0.618,  fig.show = "hold",
  out.width = "90%")
```


```{r}
library(ggplot2);
theme_set(theme_minimal())
library(knitr); 
library(tidyverse);
library(rstan);
library(tufte);
library(parallel);
library(cmdstanr);
library(posterior);
library("bayesplot");
library("ggplot2");
library("rstanarm");
set_cmdstan_path("/Users/hyunjimoon/Dropbox/20_paper/charles/code/cmdstan")
options(digits = 2);  options(htmltools.dir.version = FALSE)
parallel:::setDefaultClusterOptions(setup_strategy = "sequential")
scriptDir <- getwd()
modelDir <- file.path(scriptDir, "models")
dataDir <- file.path(scriptDir, "data")
nChains <- 4
parallel_chains <- min(nChains, detectCores())

set_Data_Model_Dir <- function(dataFile, modelName){
  scriptDir <- getwd()
  modelDir <- file.path(scriptDir, "models")
  dataDir <- file.path(scriptDir, "data")
  outDir <- file.path(scriptDir, "deliv", modelName)
  delivDir <- file.path(scriptDir, "deliv", modelName)
  dir.create(delivDir)
  data <- read_rdump(file.path(dataDir, dataFile))
  print(data)
  file <- file.path(modelDir, modelName, paste0(modelName, ".stan"))
  mod <- cmdstan_model(file)
  return(list(data = data, mod = mod, delivDir = delivDir)) 
}
set_Model_Dir <- function(modelName){
  scriptDir <- getwd()
  modelDir <- file.path(scriptDir, "models")
  dataDir <- file.path(scriptDir, "data")
  outDir <- file.path(scriptDir, "deliv", modelName)
  delivDir <- file.path(scriptDir, "deliv", modelName)
  dir.create(delivDir)
  file <- file.path(modelDir, modelName, paste0(modelName, ".stan"))
  mod <- cmdstan_model(file)
  return(list(data = data, mod = mod, delivDir = delivDir)) 
}
println <- function(msg) cat(msg); cat("\n")
printf <- function(pattern, ...) println(sprintf(pattern, ...))
util <- new.env()
source('tools/stanTools.R', local=util)
source('tools/sbcTools.R', local=util)

c_light <- c("#DCBCBC")
c_light_highlight <- c("#C79999")
c_mid <- c("#B97C7C")
c_mid_highlight <- c("#A25050")
c_dark <- c("#8F2727")
c_dark_highlight <- c("#7C0000")

c_light_trans <- c("#DCBCBC80")
c_light_highlight_trans <- c("#C7999980")
c_mid_trans <- c("#B97C7C80")
c_mid_highlight_trans <- c("#A2505080")
c_dark_trans <- c("#8F272780")
c_dark_highlight_trans <- c("#7C000080")

c_light_teal="#6B8E8E"
c_mid_teal="#487575"
c_dark_teal="#1D4F4F"

c_green_trans <- c("#00FF0080")
c_superfine <- c("#8F272705")
```
# 2. likelihood comparison
It is known that bernouli likelihood, log-concave though it is, is not friendly to approximation compared to normal, Poisson, binomial, and negative binomial likelihood. The approximation introduces a bias(Kuss and Rasmussen (2005), Vanhatalo, Pietiläinen, and Vehtari (2010), Cseke and Heskes (2011), Vehtari et al. (2016)). Therefore, the sbc of poisson likelihood shows greater uniformity than bernoulli.
  
```{r}
modelName <- "dm_bernoulli_ela_sbc"
dataName <- "disease_data_20.r"
res <- set_Data_Model_Dir(dataName, modelName)
data <- res$data
mod <- res$mod
delivDir <- res$delivDir
data$alpha_mu_prior <- 0.7
data$alpha_sd_prior <- 0.05
data$rho_mu_prior <- 0.7
data$rho_sd_prior <- 0.05
N = 500 #(bins = 20)*N # num of sim and fit
L = 297 # 897 (6n -3) # num of post. draw - related to iter_sampling - not used currently

sbc_res_me_sd1<- sbc(mod, modelName, data = data, N = N, L = L, save_progress = delivDir)
uniformity.sbc(sbc_res_me_sd1, modelName)
plot.sbc(sbc_res_me_sd1, modelName)
```


## 3. compuational method comparison
advi(or inla) both of which are known to indrouce bias is compared with ela inference. 
The alternative to SBC could be [PSIS](https://mc-stan.org/loo/reference/pareto-k-diagnostic.html) that is performed based on log_lik in generated quantities block. PSIS diagnostics might be computationally cheaper than SBC, but is a diagnostics that concentrates on specific data rather than model itself. Moreover, PSIS cannot assess the faithfulness of our computational methods.

If the rank outcome could be expressed as the function of posterior $E_{post}[I(\alpha <\alpha_{sim})]$ then, sensitivity of 

#model naming
empirical mean: no m, yes sd /e.g. disease_map_ela_sbc_n_dot1
0 mean: yes m, yes sd /e.g. disease_map_ela_sbc_n_m0_sddot1
<!-- for (i in 1:N){ -->
<!--   tmp <- sbc_res2$fits[[i]]$output_files() -->
<!--   rstanfit_sbc <- read_stan_csv(tmp) -->
<!--   util$check_all_diagnostics(rstanfit_sbc) -->
<!--   partition <- util$partition_div(rstanfit_sbc) -->
<!--   div_samples <- partition[[1]] -->
<!--   nondiv_samples <- partition[[2]] -->

<!--   par(mfrow=c(1, 3)) -->
<!--   plot(nondiv_samples$rho, nondiv_samples$alpha, log="xy", -->
<!--        col=c_dark_trans, pch=16, cex=0.8, xlab="rho", ylab="alpha") -->
<!--   points(div_samples$rho, div_samples$alpha, -->
<!--          col=c_green_trans, pch=16, cex=0.8) -->
<!-- } -->
<!-- theta도 해보는게 좋을까? alpha, rho (hp) 가 theta와 높은상관성 -->
<!-- plot(nondiv_samples$rho, nondiv_samples$sigma_error_ship, -->
<!--      col=c_dark_trans, pch=16, cex=0.8, xlab="length_GP_engine", ylab="sigma_error_ship") -->
<!-- points(div_samples$length_GP_engine, div_samples$sigma_error_ship, -->
<!--        col=c_green_trans, pch=16, cex=0.8) -->

<!-- plot(nondiv_samples$sigma_GP_engine, nondiv_samples$sigma_error_ship, -->
<!--      col=c_dark_trans, pch=16, cex=0.8, xlab="sigma_GP_engine", ylab="sigma_error_ship") -->
<!-- points(div_samples$sigma_GP_engine, div_samples$sigma_error_ship, -->
<!--        col=c_green_trans, pch=16, cex=0.8) -->
